---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## Package: conspriorityr

<!-- badges: start -->
[![R-CMD-check](https://github.com/AnthonyBasooma/conspriorityr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/AnthonyBasooma/conspriorityr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Introduction

Funding biodiversity conservation strategies are usually minimal, thus prioritizing habitats at high risk should be conducted. We developed and tested a conservation priority index (CPI) that ranks habitats to aid in prioritizing them for conservation (Basoom et al., 2022). This package contains has fucntion to compute the rarity, conservation scores for a particular habitat or several habitats.

## Installation

### Install the recent version from Github

``` r
#For windows operating system
#
install.packages('remotes)

install_github("AnthonyBasooma/conspriorityr")
library(conspriorityr)

```

``` r
#Mac
remotes::install_github("AnthonyBasooma/conspriorityr@HEAD")
library(conspriorityr)

```
## Usage

### Conservation prioritsation vs financial constraints

* Example: Prioritizing the conservation of a waterbody with conservation significance but with limited resources in the Lake Kyoga system in Uganda


```{r libraries, echo=FALSE}

library(conspriorityr) #compute priority scores
library(sf) #to handle the vector data/maps (Pebesma & Bivand 2023))
library(dplyr) #(Wickham et al ., 2023)
library(ggplot2) #(Wickham 2023)

```

* Data sources 

```{r data}

#Obtain hydrosheds basins for Africa and extract the Kyoga lakes system (Lehner et al., 2008)
#https://data.hydrosheds.org/file/hydrobasins/standard/hybas_af_lev01-12_v1c.zip
#Kyoga lakes extracted for analysis

kyoga <- st_read(dsn=system.file('extdata/hydrosheds', 'kyoga.shp', 
                                 package = 'conspriorityr'), quiet = TRUE)


#Species occurences archived in the package
data('gbif')

species_occurences <- gbif
```

* Data cleaning and merging

```{r analysis}

#Remove any NAs in the latitudes/longitudes
species_clean <- species_occurences[complete.cases(species_occurences$decimalLatitude),]

#Convert to sf data format

species_clean_sf <- species_clean %>% st_as_sf(coords = c('decimalLongitude','decimalLatitude'), crs=st_crs(4326))

#head(species_clean_sf)

#Filter out records within the kyoga lakes
species_clean_kyoga <- sf:: st_filter(species_clean_sf, kyoga) 

#clean habitat names

```

### Testing functionalites

* Rarity ranges from 0 : Present in all habitats to 1: found in only one system


```{r testing}

#Rarity of Haplochromis avium in the Kyoga system

haplo_rare <- rarity(data = species_clean_kyoga%>% st_drop_geometry(), habitat = 'waterbody',species = 'species',
              sp = 'Haplochromis squamulatus')

haplo_rare

#Conservation priority score for Kyoga main
kyogamain <- cpi_one(data = species_clean_kyoga%>% st_drop_geometry(), habitat = 'waterbody', species = 'species',area = 'surfacearea',iucn = 'iucnstatus',
                 hname = 'Lake Kyoga')
kyogamain
# #CPI for Agu

agu <- cpi_one(data = species_clean_kyoga %>% st_drop_geometry(), habitat = 'waterbody', species = 'species',area = 'surfacearea',iucn = 'iucnstatus', hname = 'Lake Agu')
agu

#CPI Agu 31.717 vs 0.38: Which signifies that in case of less resources Lake Agu should be prioritised for conservation due to number of threatened and endemic species per unit surface area

```

## Compute for many waterbodies or habitats

```{r comparemany}
cpiall <- cpi_all(data = species_clean_kyoga%>% st_drop_geometry(), habitat = 'waterbody',
             species='species', area = 'surfacearea', iucn = 'iucnstatus')


## Considering polygons for visualization

#Merge polygon data with species occurences 

species_clean_kyogapoly <- sf::st_join(kyoga, species_clean_kyoga)#%>% st_drop_geometry()


#Compute priority scores for Lake Agu

cpiagu_poly <- cpi_one(data = species_clean_kyogapoly%>% st_drop_geometry(), habitat = 'waterbody', species = 'species', area = 'surfacearea', iucn = 'iucnstatus',
                  hname = 'Lake Agu')

#Compute for all water bodies 

cpipol<- cpi_all(data = species_clean_kyogapoly%>% st_drop_geometry(), 
             habitat = 'waterbody',
             species='species', area = 'surfacearea', iucn = 'iucnstatus')
```
## Visualization

```{r visualize}

cpipol1 <- cpipol %>% rename(waterbody =sites)

cpipolmap <- merge(species_clean_kyogapoly, y=cpipol1, by="waterbody", all.x=F) %>%
  dplyr::select(cpi, waterbody)


finalresult <- unique.data.frame(cpipolmap)

#All waterbodies 

ggplot()+
  geom_sf(data= finalresult, aes(fill=cpi))+
  scale_fill_viridis_c(direction = 1)+
  geom_sf_text(data = finalresult, aes(label = waterbody), size=2)+
  labs(x='Longitude', y='Latitude', fill='Priority scores')

#Only water bodies with cpi > 5with high scores

ggplot()+
  geom_sf(data= finalresult %>% filter(cpi>5), aes(fill=cpi))+
  scale_fill_viridis_c(direction = 1)+
  geom_sf_text(data = finalresult %>% filter(cpi>5), aes(label = waterbody), size=3)+
  labs(x='Longitude', y='Latitude', fill='Priority scores')

```

## References

1. Basooma, A., Nakiyende, H., Olokotum, M., Balirwa, J. S., Nkalubo, W., Musinguzi, L., & Natugonza, V. (2022). A novel index to aid in prioritizing habitats for site‐based conservation. Ecology and Evolution, 12(3), e8762.
2. Lehner, B., Verdin, K., & Jarvis, A. (2008). New global hydrography derived from spaceborne elevation data. Eos, Transactions American Geophysical Union, 89(10), 93-94.
3. Pebesma, E., & Bivand, R. (2023). Spatial Data Science: With Applications in R (1st
  ed.). Chapman and Hall/CRC. https://doi.org/10.1201/9780429459016
4. Wickham H, François R, Henry L, Müller K, Vaughan D (2023). _dplyr: A Grammar of Data
  Manipulation_. R package version 1.1.2, <https://CRAN.R-project.org/package=dplyr>.
5. Wickham H, François R, Henry L, Müller K, Vaughan D (2023). _dplyr: A Grammar of Data
  Manipulation_. R package version 1.1.2, <https://CRAN.R-project.org/package=dplyr>.

### END


