---
title: "conspriorityr_vignette"
author: 'Anthony Basooma'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{conspriorityr_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r initialisation, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(conspriorityr)
library(sf)
```

## Usage

### Conservation prioritsation vs financial constraints

* Example: Prioritizing the conservation of a waterbody with conservation significance but with limited resources in the Lake Kyoga system in Uganda

* Data sources 

```{r data sources}

#Obtain HydroSHEDS basins for Africa and extract the Kyoga lakes system (Lehner et al., 2008)
#https://data.hydrosheds.org/file/hydrobasins/standard/hybas_af_lev01-12_v1c.zip
#Kyoga lakes extracted for analysis

kyoga <- st_read(dsn=system.file('extdata/hydrosheds', 'kyoga.shp', 
                                 package = 'conspriorityr'), quiet = TRUE)


#GBIF data archived in the package
gbifdata <- gbif

```

```{r priortiyscores}
#only selected water bodies: for only the priority scores
cpifinal <- priorityindex(data = gbifdata, 
                          habitat = 'waterbody', 
                          species = 'species', 
                          area = 'surfacearea', 
                          iucn = 'iucnstatus')
head(cpifinal)
```

```{r full output}
#only selected water bodies: return full data output
cpifull <- priorityindex(data = gbifdata, 
                          habitat = 'waterbody', 
                          species = 'species', 
                          area = 'surfacearea', 
                          iucn = 'iucnstatus',
                          full = TRUE)
head(cpifull)
```


## References

1. Basooma, A., Nakiyende, H., Olokotum, M., Balirwa, J. S., Nkalubo, W., Musinguzi, L., & Natugonza, V. (2022). A novel index to aid in prioritizing habitats for site‐based conservation. Ecology and Evolution, 12(3), e8762.
2. Lehner, B., Verdin, K., & Jarvis, A. (2008). New global hydrography derived from spaceborne elevation data. Eos, Transactions American Geophysical Union, 89(10), 93-94.
3. Pebesma, E., & Bivand, R. (2023). Spatial Data Science: With Applications in R (1st
  ed.). Chapman and Hall/CRC. https://doi.org/10.1201/9780429459016
4. Wickham H, François R, Henry L, Müller K, Vaughan D (2023). _dplyr: A Grammar of Data
  Manipulation_. R package version 1.1.2, <https://CRAN.R-project.org/package=dplyr>.
5. Wickham H, François R, Henry L, Müller K, Vaughan D (2023). _dplyr: A Grammar of Data
  Manipulation_. R package version 1.1.2, <https://CRAN.R-project.org/package=dplyr>.

### END












